---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README/README-fig-",
  out.width = "50%",
  fig.align = "center",
  table.format = "latex",
  results = "asis",
  digits = 4
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# call needed  packages 
library(dplyr)
library(ggplot2)
library(readr)
library(stringr)
library(hrbrthemes)
library(colorblindr)
library(cowplot)
library(kableExtra)
library(tinytex)

setwd("~/Dropbox/Funding applications/RStudio Internship/nfl_analysis")

#load data in better
data <- read_csv("data/nfl_2010-2017.csv") %>%                 # load data
  select(-X1)

data$team <- str_replace_all(data$team, c("JAC" = "JAX", "STL" = "LA", 
                                          "SD" = "LAC"))       # change team names if teams have moved

#condense individual data into yearly stats
indData <- data %>%
  select(-game_week) %>%                                       # remove irrelevant vars
  group_by(name, game_year, team) %>%                          # group by player and year 
  summarise_at(.vars = vars(-name, -team, -game_year, -position), 
               .funs = funs(sum), na.rm = TRUE) %>%            # sum all statistics and remove NAs
  ungroup()                                                    # ungroup

#add team variable back into individual data
indData <- data %>%
  select(name, position, team) %>%
  left_join(indData, by = c("name", "team"))

#condense team data into yearly stats
teamData <- data %>%
  filter(!is.na(team)) %>%                          # remove entries w/ NA team 
  select(-name, -position, -game_week, -rate) %>%   # remove irrelavant vars
  group_by(team, game_year) %>%                     # group by team and year
  summarise_all(funs(sum), na.rm = TRUE) %>%        # sum all statistics and remove NAs
  ungroup()                                         # ungroup
```

```{r, include = FALSE}
indData <- indData %>%                               # modify current data
  group_by(position, game_year) %>%                  # group by player and year
  mutate_at(vars(-name, -position, 
                 -game_year, -team),                 # unselect character vars
                                                     # and year var
            funs(z = ((. - mean(.)) / sd(.)))) %>%   # create z-score for all 
                                                     # variables

  ungroup() %>%                                      # ungroup
  select(name, position, game_year, contains("z"))   # restrict vars to zscores


rb <- indData %>%                                    # create dataset for RBs
  select(name, game_year, position, contains("rush")) %>% # select rushing vars
  filter(position == "RB") %>%                       # filter to RB position
  select(-position) %>%                              # remove position var
  mutate(rush_fumbles_z = rush_fumbles_z * -1) %>%   # reverse score fumbles
  ungroup() %>%                                      # clear any existing grouping
  group_by(name, game_year) %>%                      # group by player name
  mutate(totalZ = rush_att_z + rush_yds_z + rush_avg_z + rush_tds_z + 
              ifelse(is.na(rush_fumbles_z), 0, rush_fumbles_z)) %>%   # create total z score index
  select(name, game_year, totalZ) %>%                # select name, year, and totalZ
  distinct() %>%                                     # remove duplicate entries
  arrange(desc(totalZ))                              # arrange with greatest valuea at top

qb <- indData %>%                                    # create dataset for QBs
  select(name, position, game_year, contains("pass"), contains("int"), 
         contains("sck"), contains("rate")) %>%      # select passing vars
  filter(position == "QB") %>%                       # filter down to QB position
  select(-position) %>%                              # remove position var
  mutate(int_z_sumZ = int_z * -1,
         sck_z_sumZ = sck_z * -1,
         pass_fumbles_z = pass_fumbles_z * -1) %>%   # reverse score ints, scks, and fumbles
  ungroup() %>%                                      # clear any existing grouping
  group_by(name, game_year) %>%                      # group by player name
  mutate(totalZ = pass_att_z + pass_yds_z + pass_tds_z + int_z + sck_z + 
              ifelse(is.na(pass_fumbles_z), 0, pass_fumbles_z) + rate_z) %>%  # create total z score index
  select(name, game_year, totalZ) %>%                # select name, year and totalZ
  distinct() %>%                                     # remove duplicate entries
  arrange(desc(totalZ))                              # arrange with greatest valuea at top

wr_te <- indData %>%                                 # create dataset for WRs and TEs
  select(name, position, game_year, contains("rec")) %>% # select receiving vars
  filter(position == "WR/TE") %>%                    # filter down to wr/te
  select(-position) %>%                              # remove position var
  ungroup() %>%                                      # remove any existing groupings
  group_by(name, game_year) %>%                      # group by player name
  mutate(rec_fumbles_z = rec_fumbles_z * -1) %>%     # reverse score fumbles
  mutate(totalZ = rec_z + rec_yds_z + rec_avg_z + rec_tds_z + 
              ifelse(is.na(rec_fumbles_z), 0, rec_fumbles_z)) %>% # create total z score index
  select(name, game_year, totalZ) %>%                # select name, year, and totalZ
  distinct() %>%                                     # remove duplicate entries
  arrange(desc(totalZ))                              # arrange with greatest valuea at top

rbCareer <- rb %>%                                   # specify rb dataset
  ungroup() %>%                                      # remove previous groupings
  group_by(name) %>%                                 # group by name
  summarise(careerZ = sum(totalZ, na.rm = TRUE)) %>% # sum all zscores by person
  arrange(desc(careerZ))                             # arrange in a descending order
  
qbCareer <- qb %>%                                   # specify qb data
  ungroup() %>%                                      # remove previous groupings
  group_by(name) %>%                                 # group by name
  summarise(careerZ = sum(totalZ, na.rm = TRUE)) %>% # sum all zscores by person
  arrange(desc(careerZ))                             # arrange in a descending order

wr_teCareer <- wr_te %>%                             # specify wr_te data
  ungroup() %>%                                      # remove previous groupings
  group_by(name) %>%                                 # group by name
  summarise(careerZ = sum(totalZ, na.rm = TRUE)) %>% # sum all zscores by person
  arrange(desc(careerZ))                             # arrange in a descending order

rb                                                   # print rb

rbCareer                                             # print rb career data

qb                                                   # print qb

qbCareer                                             # print qb career data

wr_te                                                # print wr_te

wr_teCareer                                          # print wr_te career data
```

```{r include = FALSE}
teamData <- teamData %>%                             # modify current data
  group_by(game_year) %>%                            # group by player and year
  mutate_at(vars(-team, -game_year),                 # unselect character vars
                                                     # and year var
            funs(z = ((. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE)))) %>%   # create z-score for all 
                                                     # variables 

  ungroup() %>%                                      # ungroup
  select(team, game_year, contains("z")) %>%         # restrict vars to zscores
  group_by(team, game_year) %>%                      # group by name and position
  summarise_at(vars(-team, -game_year), funs(sumZ = sum(., na.rm = TRUE))) %>%    #sum zscores across years
  mutate(rush_fumbles_z_sumZ = rush_fumbles_z_sumZ * -1,
         rec_fumbles_z_sumZ = rec_fumbles_z_sumZ * -1,
         int_z_sumZ = int_z_sumZ * -1,
         sck_z_sumZ = sck_z_sumZ * -1,
         pass_fumbles_z_sumZ = pass_fumbles_z_sumZ * -1) %>% # reverse score fumbles, sacks, and ints
  mutate(totalZ = rush_att_z_sumZ + rush_yds_z_sumZ + rush_avg_z_sumZ + 
              rush_tds_z_sumZ + 
              ifelse(is.na(rush_fumbles_z_sumZ), 0, rush_fumbles_z_sumZ) + 
              rec_z_sumZ + 
              rec_yds_z_sumZ + rec_avg_z_sumZ + rec_tds_z_sumZ +
              rec_fumbles_z_sumZ + pass_att_z_sumZ + pass_yds_z_sumZ + 
              pass_tds_z_sumZ + int_z_sumZ + sck_z_sumZ + 
              pass_fumbles_z_sumZ) %>%               # create total z score index
  select(team, game_year, totalZ) %>%
  arrange(desc(totalZ))                              # sort in descending order

teamDataWindow <- teamData %>%                       # store teamData in teamDataWindow
  group_by(team) %>%                                 # group by team
  summarise(windowZ = sum(totalZ)) %>%               # sum z scores across team
  arrange(desc(windowZ))                             # arrange in descending order

teamData                                             # print teamData

teamDataWindow                                       # print teamDataWindow
```

# GitHub README
This project is a reproducible R project that is analyzing data collected from 
the National Football League (NFL) between the 2000 and 2017 seasons. Traditionally, most statistics in sports are compared based on the raw totals (e.g., comparing the number of career passing yards between players). The unfortunate downfall of such an approach is that the trends in the league 
differ, and as a result, players’ raw statistical totals may reflect those 
trends. For example, players who played in the early decades of the NFL were playing in a rushing dominant context, whereas players now are playing in a 
passing dominant context.

This statistical analysis will attempt to serve as a basis for comparing 
players’ statistical totals regardless of when each player played. This will be done by calculating z-scores based on the yearly statistics for all statistical categories. Subsequently, yearly totals can be calculated by summing all of the variables for each year. Then, yearly totals can be compared across years, and yearly totals can be summed to calculate career totals based off of those 
z-scores.

# Data Analysis Questions
In comparing sports statistics over long periods of time, it is often difficult 
to make direct comparisons given contextual changes in styles of gameplay over time. This data analysis seeks to use z-scores to allow for comparison across years. This data analysis will utilize z-score comparisons to demonstrate which players have had the most prolific seasons and careers. Similarly, team data 
will be analyzed to see which teams have had the most prolific seasons.  

# Data Analysis Findings
Because of the significant differences between positions, the season and career results are delineated by position. The three positions examined are 
quarterback, running back, and receiving positions (wide receivers and tight 
ends). The team season statistics are an amalgamation of the performances from these three positions.

## Running backs

```{r}
# graph individuals' season data
rb %>%
  ggplot(aes(x = game_year, y = totalZ)) +                                   # set vars for axis
  geom_jitter(alpha = 5/9, color = "green") +                                # set alpha and color
  theme(panel.background = element_rect(fill = 'black', colour = 'black'),   # set background to black
        panel.grid.major = element_blank(),                                  # remove major grid lines
        panel.grid.minor = element_blank()) +                                # remove minor grid lines
  ggtitle("RB Performance by Season") +                                      # set graph title
  labs(x = "Year", y = "Total Z Score")                                      # set axes labels
```

### Running back season performances

```{r}
kable(head(rb, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = "A table of the top 30 running back performances by season.")
```

### Running back career performances

```{r}
kable(head(rbCareer, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = "A table of the top 30 running backs by career.")
```

According to this analysis, Shaun Alexander's performance from 2005 is the best running back single season performance within the analysis window. As can be 
seen from the graph as well as the table of season performances, there are many single season performances that are ranked closely behind. In regards to the 
most prolific running back career within the analysis window, LeDainian 
Tomlinson is clearly above his peers in terms of career performance. That being said, LeSean McCoy and Frank Gore also appear to be outliers in terms of career performance in relation to their peers.

## Quarterbacks

```{r}
qb %>%
  ggplot(aes(x = game_year, y = totalZ)) +                                   # set vars for axis
  geom_jitter(alpha = 5/9, color = "green") +                                # set alpha and color
  theme(panel.background = element_rect(fill = 'black', colour = 'black'),   # set background to black
        panel.grid.major = element_blank(),                                  # remove major grid lines
        panel.grid.minor = element_blank()) +                                # remove minor grid lines
  ggtitle("QB Performance by Season") +                                      # set graph title
  labs(x = "Year", y = "Total Z Score")                                      # set axes labels
```

### Quarterback season performances
```{r}
kable(head(qb, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = 'A table of the top 30 quarter back performances by season.')
```

### Quarterback career performances
```
kable(head(qbCareer, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = 'A table of the top 30 quarter backs by career.')
```

According to this analysis, Drew Bledsoe's performance from 2002 is the best quarterback single season performance within the analysis window. Similar to the situation with the running backs, there are numerous other quarterbacks with similar single season performances. In regards to the most prolific quarterback career within the analysis window, Peyton Manning has the highest career rating amongst the quarterbacks. Once again, there appear to be several other 
quarterbacks who are outliers amongst their peers. These other outliers include notable quarterbacks such as Drew Brees, Brett Favre, Tom Brady, and Ben Roethlisberger.

## Receiving

```{r}
wr_te %>%
  ggplot(aes(x = game_year, y = totalZ)) +                                   # set vars for axis
  geom_jitter(alpha = 5/9, color = "green") +                                # set alpha and color
  theme(panel.background = element_rect(fill = 'black', colour = 'black'),   # set background to black
        panel.grid.major = element_blank(),                                  # remove major grid lines
        panel.grid.minor = element_blank()) +                                # remove minor grid lines
  ggtitle("WR/TE Performance by Season") +                                   # set graph title
  labs(x = "Year", y = "Total Z Score")                                      # set axes labels
```

### Receiving season performances

```{r}
kable(head(wr_te, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = 'A table of the top 30 receiving performances by season.')
```

### Receiving career performances

```{r}
kable(head(wr_teCareer, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = 'A table of the top 30 receviers and tight ends by career.')
```

According to this analysis, Randy Moss' performance from 2007 is the best 
receiving single season performance within the analysis window. In regards to 
the most prolific receiving career within the analysis window, Terrell Owens narrowly claimed the top honor ahead of Tony Gonzalez. Unlike the running back 
and quarterback categories, there does not appear to be the same clear differentiation between several the elite players and the very  good players.

## Team

```{r}
# graph team data
teamData %>%
  ggplot(aes(x = game_year, y = totalZ, color = team)) +                     # set vars for axis
  geom_jitter(alpha = 5/9) +                                                 # set alpha
  theme(panel.background = element_rect(fill = 'black', colour = 'black'),   # set background to black
        panel.grid.major = element_blank(),                                  # remove major grid lines
        panel.grid.minor = element_blank()) +                                # remove minor grid lines
  ggtitle("Team Performance by Season") +                                    # set graph title
  labs(x = "Year", y = "Total Z Score")                                      # set axes labels
```

### Team offensive season performances

```{r}
kable(head(teamData, 30), booktabs = TRUE, format = 'markdown', align = 'c',
        caption = 'A table of the top 30 offensive team performances by season.')
```

According to this analysis, the best team offensive performance was done by the 
New Orleans Saints in 2011.  Other notable offensive team performances were achieved by the New England Patriots (2007, 2012), and Denver (2013).

Of note, the interpretations of these findings, especially the findings 
regarding individual players, must be interpreted with caution given the limited window of analysis for this data project. The window of analysis only extends 
from 2000 through 2017. This means that players whose entire careers are not contained within this window may have been underestimated. Furthermore, the majority of players in the NFL's history have played outside of this window. 
Thus, this is a limited view into the best statistical seasons and careers.